---
title: "My Final Project Template"
author: Madalyn Benson
subtitle: Predictability of Harmful Algal Bloom Prevalance in NYS
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

An algal bloom is the overgrowth of algae in water sources. Algal blooms
occur when exposed to the correct environmental conditions (i.e. excess
nutrients, warm temperatures, ample sunlight, and low-flow water
conditions). A bloom can be short-lived (a couple of hours) or
long-lived (a couple of weeks) depending on how long the algae persist
in the environment. Harmful algal blooms (HABs) can be harmful to humans
and the environment. When this occurs the algal species can produce
toxins with serious health impacts to humans, degrade aquatic
ecosystems, and also affect the economy near coastal areas. The dominant
algae in New York State include but are not limited to: Green algae
(Chlorophyta) Golden brown algae (Chrysophyta), which also includes a
large group of diatoms Blue-green algae (Cyanobacteria) Several smaller
groups including (euglenoids, cryptomonads, and dinoflagellates)

The presence of the above harmful algal blooms varies both spatially and
temporally in NYS, which results in varying risk for people, animals,
and the local ecology. HABs are most prevalent in the months June
through October, however they can occur at any time of year. HABs reduce
tourism, close beaches, and greatly diminish local economies. Because of
this unpredictability, harmful algal blooms are of national concern.
Data suggests that certain environmental conditions can be used as a
tool to better understand and predict HABs. This project aims to explore
any trends between harmful algal blooms prevalence within New York State
by relating HAB prevalence over time to environmental conditions and
populations.

# Materials and methods

## Example 1

![Water Quality Assessment of New York
State](https://www.dec.ny.gov/images/water_images/305basmtmap10.jpg)

Image 1 is the extracted raster data of the New York State water quality
report from the New York State Department of Environmental Conservation
(DEC). I selected this image as an example because I will be needing a
raster data image of New York for my project. This is a useful resource
because it shows the water bodies in NYS. In my final project I will be
showing where and when harmful algal blooms affect the water bodies of
New York. <https://www.dec.ny.gov/chemical/66532.html>

## Example 2

![Climate Data
Grapher](https://www.nyclimatescience.org/static/media/datagrapher.14b4fecf.png)

Image 2 is an interactive map of annual maximum temperatures in New York
State from the New York Climate Change Science Clearinghouse (NYCCSC).
This is a valuable data source because the user can select data from New
York State stations from gridded observations, spatially aggregated by
New York counties and river basins, or by states neighboring New York.
Different time periods can be observed to see temperature changes over
time. <https://www.nyclimatescience.org/highlights/data_products>

## Example 3

![Freqeuncy of HAB
Events](https://www.researchgate.net/profile/Kb-Padmakumar/publication/236142114/figure/fig4/AS:328425987428355@1455314265478/Frequency-of-occurrence-of-harmful-algal-blooms-during-the-last-century.png)

Image 3 is a figure displaying the occurrence of HAB events over time.
This is a useful graphic because it is showing a sharp increase in algal
blooms over 100 years. I would like to create a chart similar to this
inorder to provide reasoning for an increase in algal blooms across NYS.
I would do this by comparing other correlating data trends to support an
increase in algal blooms.
<https://www.researchgate.net/figure/Frequency-of-occurrence-of-harmful-algal-blooms-during-the-last-century_fig4_236142114>

# Proposed data sources

1.  Water quality data <https://www.dec.ny.gov/chemical/31290.html> To
    get water quality data for NYS, I plan on using the NYS Section
    303(d) list of impaired/TMDL waters. Every two years, the Federal
    Clean Water Act and supporting federal regulations require states to
    submit a list of impaired waters that require the development of a
    Total Maximum Daily Load (TMDL). I will be using the 2020-2022 Clean
    Water Act (CWA) data set.

2.  Spatial polygon data of NYS Counties
    <http://gis.ny.gov/gisdata/metadata/nysgis.Counties_Shoreline.pdf>
    To get polygon data for NYS, I plan on using the NYS Civil
    Boundaries County Shoreline data from NYS GIS Clearinghouse. This
    dataset has accurate polygon edges for the edge of NYS counties.
    This is essential because these polygons will be used to extract
    data from the NLCD dataset, which has accurate data on the
    shorelines of state and county polygons.

3.  Spatial polygon data of algal blooms
    <https://www.dec.ny.gov/chemical/83332.html> In order to track the
    HABs in NYS, I will be using the DEC 2012-2019 HABs archive summary
    to summarize all water bodies with harmful algal blooms documented.
    This will show algal bloom trends in NYS over time.

4.  Spatial data showing NYS temperatures over time
    <https://www.nyclimatescience.org/highlights/data_products> I will
    be using the interactive map of annual maximum temperatures in New
    York State from the New York Climate Change Science Clearinghouse
    (NYCCSC) to select data from New York State stations and observe
    temperature trends over time. I could also use NOAA temperatures at
    varying weather stations across NYS.

5.  Annual Population Estimates for New York State and Counties:
    Beginning 1970
    <https://data.ny.gov/Government-Finance/Annual-Population-Estimates-for-New-York-State-and/krt9-ym2k/data>
    I will be using this data to correlate a relationship between
    population density and HAB prevalence.

# Proposed methods

I will use the raster or read_sf package to download spatial polygon
data for NYS counties. I will use the merge function to join all data
and the tmap package to create the choropleth maps. To create an
interactive map to display the data listed above, I will use the leaflet
package.

1.  creating a choropleth map for: average populations at county level
    across NYS

2.  creating a time-series graph of temperature trends over time for NYS
    counties

3.  creating interactive maps that display the following when a country
    is clicked on: county, year, and type of algal bloom

# Expected results

Overall, I believe this project will be a successful attempt at
monitoring the predictability of harmful algal bloom prevalence in NYS.
Based on my expectations of this outline, relevant data will be
projected in an effective way to show statistically relevant
correlations.

# Materials and Methods
### Create a choroplethr map of average populations at county level across NYS
```{r}
library(choroplethr)
library(choroplethrMaps)
library(janitor)
#default parameters
data(df_pop_county)
#zoom into new york state with additional reference map
county_choropleth(df_pop_county,
                  title = "New York County Population Estimates, 2012",
                  legend = "Population",
                  state_zoom = "new york")
```
### Create a time-series graph of temperature trends over time for weather stations across NYS
```{r}
library(tidyverse)
library(dplyr)
library(p8105.datasets)
library(patchwork)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

data(ny_noaa)
head(ny_noaa, 10)

ny_noaa_structure = print(str(ny_noaa))

ny_noaa_tidy = 
  ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, into = c("Year", "Month", "Day")) %>%
  mutate(tmax = as.numeric(tmax), tmin = as.numeric(tmin), Year = as.numeric(Year), Month = as.numeric(Month), Day = as.numeric(Day)) %>%
   mutate(Month = month.name[Month]) %>%
  mutate(tmax = tmax/10, 
         tmin = tmin/10, 
         prcp = prcp/10)

ny_noaa_summary =
  ny_noaa_tidy %>%
  filter(Month %in% c("January", "July")) %>%
  group_by(id, Year, Month) %>%
  summarize(mean_tmax = mean(tmax), na.rm = TRUE)

ggplot(ny_noaa_summary, aes(x = Year, y = mean_tmax, group = Year, color = Month)) +
    geom_boxplot() + 
    facet_grid(. ~ Month) +
    theme(legend.position = "none") + 
  labs(title = "Average Maximum Temperature for New York State Weather Stations",y = "Average Maximum Temperature (C)")

```
### Create an interactive map displaying county, year, and type of algal bloom when a waterbody is clicked on
```{r}
library(ggplot2)
library(maps)
library(leaflet)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tmap)

noaa_df = readr::read_csv("data/Harmful_Algal_Bloom_Statewide_Occurrence_Summary__2012-2018.csv")
 
str(noaa_df)
m <- leaflet(noaa_df, options = leafletOptions(minZoom = 6, dragging = TRUE)) %>%
  setView(lat = 43, lng = -75, zoom = 6) %>%
  setMaxBounds(lng1 = -70, 
               lat1 = 40, 
               lng2 = -80, 
               lat2 = 45) %>%
  addTiles() %>%
  addMarkers(data = noaa_df,
             popup= ~paste(`Waterbody Name`, "<br>County:",`County`, "<br>Bloom Type:", `Bloom Type`,"<br>Coordinates:", `Location 1`))
m  # Print the map

```

# Results and Conclusions

Algal blooms occur when exposed to the correct environmental conditions (i.e. excess nutrients, warm temperatures, ample sunlight, and low-flow water conditions). The maps and other graphics created above are carefully planned to
convey the results of harmful algal blooms increasing over time in New York State. A map of New York State counties is provided as a reference map to refer back to in regards to populations. As shown on the map, the darker counties have larger populations. Downstate counties towards New York City and Long Island have some of the highest populations per county. This corresponds with the leaflet map because there are higher amounts of algal bloom occurrences in these higher populated counties. Higher populated areas tend to have more impaired water bodies and can be a great source nutrients for algae to grow.

The average maximum temperatures for New York State weather stations was not the results I expected. From 1980-2010 the maximum temperatures did not fluctuate too much. I was predicting there to be a higher increase in temperature maximums due to global warming and an increased amount of algal blooms. Higher temperatures typically lead to more algae blooms. Further investigation to include more recent weather station data could possibly change the data to have more expected results. 


# References

American Community Survey 5-Year Data (2009-2021). (2022, December 9). Census Bureau. Retrieved December 16, 2022, from https://www.census.gov/data/developers/data-sets/acs-5year.html

Datasets | Climate Data Online (CDO) | National Climatic Data Center (NCDC). (n.d.). National Centers for Environmental Information. Retrieved December 16, 2022, from https://www.ncdc.noaa.gov/cdo-web/datasets

Harmful Algal Blooms (HABs) - NYS Dept. of Environmental Conservation. (n.d.). New York State Department of Environmental Conservation. Retrieved December 16, 2022, from https://www.dec.ny.gov/chemical/77118.html

