---
title: "My Final Project Template"
author: Madalyn Benson
subtitle: Predictability of Harmful Algal Bloom Prevalance in NYS
---

# Introduction

An algal bloom is the overgrowth of algae in water sources. Algal blooms occur when exposed to the correct environmental conditions (i.e. excess nutrients, warm temperatures, ample sunlight, and low-flow water conditions). A bloom can be short-lived (a couple of hours) or long-lived (a couple of weeks) depending on how long the algae persist in the environment. Harmful algal blooms (HABs) can be harmful to humans and the environment. When this occurs the algal species can produce toxins with serious health impacts to humans, degrade aquatic ecosystems, and also affect the economy near coastal areas. The dominant algae in New York State include but are not limited to:
  Green algae (Chlorophyta)
  Golden brown algae (Chrysophyta), which also includes a large group of diatoms
  Blue-green algae (Cyanobacteria)
  Several smaller groups including (euglenoids, cryptomonads, and dinoflagellates)
  
The presence of the above harmful algal blooms varies both spatially and temporally in NYS, which results in varying risk for people, animals, and the local ecology. HABs are most prevalent in the months June through October, however they can occur at any time of year. HABs reduce tourism, close beaches, and greatly diminish local economies.  Because of this unpredictability, harmful algal blooms are of national concern. Data suggests that certain environmental conditions can be used as a tool to better understand and predict HABs. This project aims to explore any trends between harmful algal blooms prevalence within New York State by relating HAB prevalence over time to environmental conditions and populations. 

# Materials and methods

[~ 200 words]

Narrative: Clear narrative description of the data sources and methods. Includes data from at least two sources that were integrated / merged in R.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. If you want to use your own data, you must make it available on a website (e.g. Figshare) so that others are able to re-run your code.

You can do bullets like this:

* The first most important thing
* The second most important thing
* The third most important thing

You can do numbers like this:

1. The first most important thing
2. The second most important thing
3. The third most important thing

See [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/) for all the amazing things you can do.


Here's my first code chunk.
```{r}
1+2
```

Load any required packages in a code chunk (you may need to install some packages):

```{r, message=F, warning=F}
library(tidyverse)
library(leaflet)
library(kableExtra)
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Download and clean all required data
```{r}
n=20
data=data.frame(x=runif(n,-180,180),
                y=runif(n,-60,60),
                size = runif(n, 5, 20),
                category = factor(
                  sample(letters[1:5], n, replace = TRUE)
                  ),
                value = rnorm(n))
```

```{r, results='asis'}
data %>% 
  slice(1:10) %>% #show only 1:n rows
  kable(digits=2,align="c")%>% #make table and round to two digits
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive")) #apply other formatting
```

Add any additional processing steps here.

# Results

[~200 words]

Tables and figures (maps and other graphics) are carefully planned to convey the results of your analysis. Intense exploration and evidence of many trials and failures. The author looked at the data in many different ways before coming to the final presentation of the data.

Show tables, plots, etc. and describe them.

```{r, fig.width=6, fig.height=3, fig.cap="Map of completely random data"}
m <- leaflet(data) %>% 
  addTiles() %>% 
  addCircleMarkers(~x, ~y, radius = ~size,color = ~as.factor(category)) %>% 
  addPopups(~x[2], ~y[2], "Random popup")
m  # a map with the default OSM tile layer
```


```{r}
data %>% 
  ggplot(aes(x=x,y=y,col=category))+
  geom_point()
```

# Conclusions

[~200 words]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner
