---
title: "My Final Project Template"
author: Madalyn Benson
subtitle: Predictability of Harmful Algal Bloom Prevalance in NYS
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

An algal bloom is the overgrowth of algae in water sources. Algal blooms
occur when exposed to the correct environmental conditions (i.e. excess
nutrients, warm temperatures, ample sunlight, and low-flow water
conditions). A bloom can be short-lived (a couple of hours) or
long-lived (a couple of weeks) depending on how long the algae persist
in the environment. Harmful algal blooms (HABs) can be harmful to humans
and the environment. When this occurs the algal species can produce
toxins with serious health impacts to humans, degrade aquatic
ecosystems, and also affect the economy near coastal areas. The dominant
algae in New York State include but are not limited to: Green algae
(Chlorophyta) Golden brown algae (Chrysophyta), which also includes a
large group of diatoms Blue-green algae (Cyanobacteria) Several smaller
groups including (euglenoids, cryptomonads, and dinoflagellates)

The presence of the above harmful algal blooms varies both spatially and
temporally in NYS, which results in varying risk for people, animals,
and the local ecology. HABs are most prevalent in the months June
through October, however they can occur at any time of year. HABs reduce
tourism, close beaches, and greatly diminish local economies. Because of
this unpredictability, harmful algal blooms are of national concern.
Data suggests that certain environmental conditions can be used as a
tool to better understand and predict HABs. This project aims to explore
any trends between harmful algal blooms prevalence within New York State
by relating HAB prevalence over time to environmental conditions and
populations.

# Materials and methods

## Example 1

![Water Quality Assessment of New York
State](https://www.dec.ny.gov/images/water_images/305basmtmap10.jpg)

Image 1 is the extracted raster data of the New York State water quality
report from the New York State Department of Environmental Conservation
(DEC). I selected this image as an example because I will be needing a
raster data image of New York for my project. This is a useful resource
because it shows the water bodies in NYS. In my final project I will be
showing where and when harmful algal blooms affect the water bodies of
New York. <https://www.dec.ny.gov/chemical/66532.html>

## Example 2

![Climate Data
Grapher](https://www.nyclimatescience.org/static/media/datagrapher.14b4fecf.png)

Image 2 is an interactive map of annual maximum temperatures in New York
State from the New York Climate Change Science Clearinghouse (NYCCSC).
This is a valuable data source because the user can select data from New
York State stations from gridded observations, spatially aggregated by
New York counties and river basins, or by states neighboring New York.
Different time periods can be observed to see temperature changes over
time. <https://www.nyclimatescience.org/highlights/data_products>

## Example 3

![Freqeuncy of HAB
Events](https://www.researchgate.net/profile/Kb-Padmakumar/publication/236142114/figure/fig4/AS:328425987428355@1455314265478/Frequency-of-occurrence-of-harmful-algal-blooms-during-the-last-century.png)

Image 3 is a figure displaying the occurrence of HAB events over time.
This is a useful graphic because it is showing a sharp increase in algal
blooms over 100 years. I would like to create a chart similar to this
inorder to provide reasoning for an increase in algal blooms across NYS.
I would do this by comparing other correlating data trends to support an
increase in algal blooms.
<https://www.researchgate.net/figure/Frequency-of-occurrence-of-harmful-algal-blooms-during-the-last-century_fig4_236142114>

# Proposed data sources

1.  Water quality data <https://www.dec.ny.gov/chemical/31290.html> To
    get water quality data for NYS, I plan on using the NYS Section
    303(d) list of impaired/TMDL waters. Every two years, the Federal
    Clean Water Act and supporting federal regulations require states to
    submit a list of impaired waters that require the development of a
    Total Maximum Daily Load (TMDL). I will be using the 2020-2022 Clean
    Water Act (CWA) data set.

2.  Spatial polygon data of NYS Counties
    <http://gis.ny.gov/gisdata/metadata/nysgis.Counties_Shoreline.pdf>
    To get polygon data for NYS, I plan on using the NYS Civil
    Boundaries County Shoreline data from NYS GIS Clearinghouse. This
    dataset has accurate polygon edges for the edge of NYS counties.
    This is essential because these polygons will be used to extract
    data from the NLCD dataset, which has accurate data on the
    shorelines of state and county polygons.

3.  Spatial polygon data of algal blooms
    <https://www.dec.ny.gov/chemical/83332.html> In order to track the
    HABs in NYS, I will be using the DEC 2012-2019 HABs archive summary
    to summarize all water bodies with harmful algal blooms documented.
    This will show algal bloom trends in NYS over time.

4.  Spatial data showing NYS temperatures over time
    <https://www.nyclimatescience.org/highlights/data_products> I will
    be using the interactive map of annual maximum temperatures in New
    York State from the New York Climate Change Science Clearinghouse
    (NYCCSC) to select data from New York State stations and observe
    temperature trends over time. I could also use NOAA temperatures at
    varying weather stations across NYS.

5.  Annual Population Estimates for New York State and Counties:
    Beginning 1970
    <https://data.ny.gov/Government-Finance/Annual-Population-Estimates-for-New-York-State-and/krt9-ym2k/data>
    I will be using this data to correlate a relationship between
    population density and HAB prevalence.

# Proposed methods

I will use the raster or read_sf package to download spatial polygon
data for NYS counties. I will use the merge function to join all data
and the tmap package to create the choropleth maps. To create an
interactive map to display the data listed above, I will use the leaflet
package.

1.  creating a choropleth map for: average populations at country level
    across NYS

2.  creating a time-series graph of temperature trends over time for NYS
    counties

3.  creating a map of NYS impaired water bodies

4.  creating interactive maps that display the following when a country
    is clicked on: population HAB occurrences

# Expected results

Overall, I believe this project will be a successful attempt at
monitoring the predictability of harmful algal bloom prevalence in NYS.
Based on my expectations of this outline, relevant data will be
projected in an effective way to show statistically relevalant
correlations.

Load any required packages in a code chunk (you may need to install some
packages):

## Download and clean all required data

```{r}
library(choroplethr)
library(choroplethrMaps)
library(janitor)
#default parameters
data(df_pop_county)
#zoom into new york state with additional reference map
county_choropleth(df_pop_county,
                  title = "New York County Population Estimates, 2012",
                  legend = "Population",
                  state_zoom = "new york")
```

```{r}
library(tidyverse)
library(dplyr)
library(p8105.datasets)
library(patchwork)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

data(ny_noaa)
head(ny_noaa, 10)

ny_noaa_structure = print(str(ny_noaa))

ny_noaa_tidy = 
  ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, into = c("Year", "Month", "Day")) %>%
  mutate(tmax = as.numeric(tmax), tmin = as.numeric(tmin), Year = as.numeric(Year), Month = as.numeric(Month), Day = as.numeric(Day), snow = as.numeric(snow)) %>%
   mutate(Month = month.name[Month]) %>%
  mutate(tmax = tmax/10, 
         tmin = tmin/10, 
         prcp = prcp/10)

ny_noaa_summary =
  ny_noaa_tidy %>%
  filter(Month %in% c("January", "July")) %>%
  group_by(id, Year, Month) %>%
  summarize(mean_tmax = mean(tmax), na.rm = TRUE)

ggplot(ny_noaa_summary, aes(x = Year, y = mean_tmax, group = Year, color = Month)) +
    geom_boxplot() + 
    facet_grid(. ~ Month) +
    theme(legend.position = "none") + 
  labs(title = "Average Maximum Temperature for New York State Weather Stations",y = "Average Maximum Temperature (C)")

```

```{r}
library(shiny)
library(shinythemes)
library(ggplot2)
library(maps)
library(leaflet)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tmap)

noaa_df = readr::read_csv("data/Harmful_Algal_Bloom_Statewide_Occurrence_Summary__2012-2018.csv")
 

m <- leaflet(noaa_df, options = leafletOptions(minZoom = 6, dragging = TRUE)) %>%
  setView(lat = 43, lng = -75, zoom = 6) %>%
  setMaxBounds(lng1 = -70, 
               lat1 = 40, 
               lng2 = -80, 
               lat2 = 45) %>%
  addTiles() %>%
  addCircles(popup= noaa_df$`County`) 
m  # Print the map
```

# Results

[\~200 words]

Tables and figures (maps and other graphics) are carefully planned to
convey the results of your analysis. Intense exploration and evidence of
many trials and failures. The author looked at the data in many
different ways before coming to the final presentation of the data.

Show tables, plots, etc. and describe them.

```{r, fig.width=6, fig.height=3, fig.cap="Map of completely random data"}

```

```{r}

```

# Conclusions

[\~200 words]

Clear summary adequately describing the results and putting them in
context. Discussion of further questions and ways to continue
investigation.

# References

All sources are cited in a consistent manner
